A Horrocks-type theorem for even orthogonal $K_2$

### 2019-09-25 dl-paper.tex
Minor

\newcommand{\XX}{\mathcal{X}}           % (subset of) Steinberg generators
\newcommand{\RR}[1]{\mathcal{R}_{#1}}   % Steinberg relations
\newcommand{\catname}[1]{{\normalfont\textbf{#1}}} %Category name

@@

For an additive subgroup $A\subseteq R$ and $\alpha \in \Phi$ we denote by $X_\alpha(A)$ the corresponding {\it root subgroup} of $\St(\Phi, R)$, i.\,e. the subgroup generated by all $x_\alpha(a)$, $a \in A$.
We denote by $U(\Phi^+, A)$ (resp. $U(\Phi^-, A)$) the subgroup of $\St(\Phi, R)$ generated by root subgroups $X_\alpha(A)$ corresponding to all positive (resp. negative) roots of $\Phi$.

@@

\subsection{Generators of relative Steinberg groups}
Denote by $\St(\Phi, I)$ the subgroup of $\St(\Phi, R)$ generated as a group by root unipotents of level $I$.
It is clear that $\overline{\St}(\Phi, R, I)$ contains $\St(\Phi, I)$ and, in fact, is its normal closure.
We also denote by $\overline{H}(\Phi, R, I)$ the subgroup of $\overline{\St}(\Phi, R, I)$ generated by the semisimple root elements $h_\alpha(u)$ and symbols $\{u, v\}$, $u \in (1+I)^\times$, $v \in R^\times$, $\alpha\in \Phi$.

@@

Recall that a closed subset $S \subseteq \Phi$ is called {\it parabolic} if $S \cup -S = \Phi$.
Its {\it special part} $\Sigma_S$, by definition, consists of all $\alpha \in S$ such that $-\alpha \not\in S$.
\begin{externaltheorem}[Stepanov] \label{thm:Stepanov} 
Let $S \subseteq \Phi$ be a parabolic subset of $\Phi$. Then the group $\overline{\St}(\Phi, R, I)$ is generated by the set $\mathcal{Z}(\Sigma_S, R, I)$.
 \end{externaltheorem} \begin{proof} See~\cite[Lemma~4]{S15}. \end{proof}

@@

\begin{lemma} \label{Crels}
The elements $c_\alpha(s, t)$ satisfy the following relations for all $s\in I,\ t\in J,\ \xi\in R$.
 \begin{enumerate}
 \item \label{C1} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = x_{\alpha}(- st\xi) \cdot x_{\alpha+\beta}(N_{\alpha,\beta}\cdot s^2t\xi)$ if $\alpha+\beta \in \Phi$;
 \item \label{C2} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = x_{\alpha}(st\xi + s^2t^2\xi) \cdot x_{\alpha-\beta}(N_{-\alpha, \beta}\cdot st^2\xi)$ if $\alpha-\beta \in \Phi$;  
 \item \label{C3} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = 1$ if $\alpha \perp \beta$;  
 \item \label{C4} If $\alpha+\beta\in\Phi$ then holds:
  \[c_{\alpha+\beta}(s, t\xi) = [x_{\beta}(st),\ x_{-\beta}(\xi)] ^ {x_{\alpha+\beta}(-s) x_{-\alpha}(\epsilon t)} \cdot c_{\alpha}(\epsilon s\xi, -\epsilon t)^{-1} \cdot x_{-\beta}(-st\xi^2),\]
  where $\epsilon = N_{\alpha,\beta}$.
 \end{enumerate}
\end{lemma}
\begin{proof}
First of all, notice that the following equalities hold for elements $x,y,z$ of any group:
\begin{align}
 [x, yz]^y =& [y^{-1}, x] \cdot [x, z], \label{rel43} \\
  [[x, y], z] =& \left([x^{-1}, [ y^{-1}, z]] ^ {y^{-1}} \cdot [y, [ z^{-1}, x^{-1}]] ^ {z^{-1}} \right)^{x^{-1}}. \label{HW-variant}
\end{align}

The first assertion follows from the following computation:
\begin{multline} \nonumber [c_{\beta}(s, t),\ x_{\alpha}(\xi)]  = [x_{\beta}(s),\ x_{-\beta}(t)] \cdot [x_{-\beta}(t),\ x_{\beta}(s)\cdot x_{\alpha + \beta}(N_{\alpha, \beta} \cdot s\xi)] = \\
 = {}^{x_\beta(s)}\![x_{-\beta}(t),\ x_{\alpha+\beta}(N_{\alpha, \beta} \cdot s\xi)] = x_{\alpha}(- st\xi) \cdot x_{\alpha+\beta}(N_{\alpha,\beta} \cdot s^2t\xi). \end{multline}

@@

 Now each factor $z_{\alpha_i}(f_i(0), \xi_i(X))$ can be written as follows:
 \[z_{\alpha_i}(f_i(0), \xi_i(0))^{x_{-\alpha_i}(X\xi'_i(X))} = [x_{-\alpha_i}(-X\xi'_i(X)),\ z_{\alpha_i}(f_i(0), \xi_i(0))] \cdot z_{\alpha_i}(f_i(0), \xi_i(0)).\]
 Since the subgroup $C_0 = \left[\St(\Phi, XA[X]), \overline{\St}(\Phi, A, M)\right]$ is normalized by $\overline{\St}(\Phi, A, M)$ 
 (this follows e.\,g. from the formula $[g, h]^{h_1} = [h_1^{-1}, g][g, h_1^{-1}h]$) we conclude that $g_1(X)$ is congruent to $\prod_i z_{\alpha_i}(f_i(0), \xi_i(0)) = g(0) = 1$ modulo $C_0$,
 which implies the assertion. \qedhere \end{proof}

\begin{externaltheorem}[Stein] \label{thm:Stein} One has $\overline{\St}(\Phi, A, M) = U(\Phi^+, M) \cdot \overline{H}(\Phi, A, M) \cdot U(\Phi^-, M).$ \end{externaltheorem} \begin{proof} See~\cite[Theorem~2.4]{Ste73}. \end{proof}

\begin{prop} \label{Kgen} The subgroup $K(A[X], M[X])$ is generated as an abstract group by the subgroup $\overline{\St}(\Phi, A[X], XM[X])$ and
 the elements $[x_\alpha(m), x_{-\alpha}(X\xi)]$, $m \in M$, $\xi \in A[X]$, $\alpha \in \Phi$. \end{prop}
\begin{proof} From~\eqref{eq:conj-h-x} we obtain that $\overline{H}(\Phi, A, M)$ normalizes both $\St(\Phi, XA[X])$ and $\St(\Phi, M)$ and, moreover, that $[\overline{H}(\Phi, A, M),\ \St(\Phi, XA[X])] \subseteq \overline{\St}(\Phi, A[X], XM[X])$.

Denote by $C_1$ the commutator subgroup $[\St(\Phi, XA[X]),\ \St(\Phi, M)]$.
It is clear that for $g \in \St(\Phi, XA[X])$, $h \in \overline{H}(\Phi, A, M)$, $u^+ \in U(\Phi^+, M)$, $u^- \in U(\Phi^-, M)$ holds:
\[ [g,\ h u^+ u^-] = [g,\ h] \cdot [{}^{h}\!g,\ {}^{h}\!(u^+u^-)] \in \St(\Phi, A[X], XM[X]) \cdot C_1.\]
Since $C_0$ is generated by the above commutators and $\overline{\St}(\Phi, A[X], XM[X])$ is a normal subgroup of $\St(\Phi, A[X])$
we obtain that $C_0 \subseteq \overline{\St}(\Phi, A[X], XM[X]) \cdot C_1$ and consequently that
$K(A[X], M[X]) = \overline{\St}(\Phi, A[X], XM[X]) \cdot C_1.$

@@

\section{Proof of the main result}
The main result of this section is~\cref{thm:P1glueing}, which is a direct generalization of~\cite[Proposition~4.3]{Tu83}. 
The object playing a key role in its proof is a certain action of the group $G = \St(\Phi, A[X\inv] + M[X])$ on a certain set $\overline{V}$, which is defined in~\cref{sec:V-construction}. Later, we will see that $\overline{V}$ is, in fact, a set-theoretic $G$-torsor.
To be able to write an explicit formula for this action we need two major ingredients. The first one is ~\cref{lemma33}, which delivers a presentation of $G$ with much fewer generations and relations than~\eqref{rel:add}-\eqref{rel:CCF}. The other ingredients are certain subgroups $P_\alpha(0)$, $P_\alpha(*)$ of $\St(\Phi, A[X, X\inv])$, which are modeled after the nameless groups from~\cite[Lemma~3.4]{Tu83}. The definition and properties of these groups are given in Sections~\ref{sec:Pa0-basic}--\ref{sec:Pa0-Steinberg}.

@@

Whenever the coefficient $\xi$ of a Steinberg generator $g = x_\alpha(\xi)$ of $\St(\Phi, B)$ is a homogeneous element of $B$, i.\,e. $\xi \in B_d$ for some $d \in \mathbb{Z}$,
 we call the corresponding generator $g$ {\it homogeneous of degree $d$}.
It is not hard to show that $\St(\Phi, B)$ can be presented by the set of all homogeneous Steinberg generators modulo the following set of Steinberg relations
 between them (below $a, a' \in B_d$, $b\in R_e$ and $d,e \in \mathbb{Z}$): 
\begin{align}
&\,\,x_{\alpha}(a)\cdot x_{\alpha}(a') =  x_{\alpha}(a+a'),                        & \tag{R$1_d$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= x_{\alpha+\beta}(N_{\alpha, \beta} \cdot ab),   & \alpha + \beta \in \Phi,\ \tag{R$2_{d,e}$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= 1,                                              & \alpha - \beta \in \Phi,\ \tag{R$3^\angle_{d,e}$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= 1,                                              & \alpha \perp \beta.\ \tag{R$3^\bot_{d,e}$}
\end{align}
By the degree of a Steinberg relation we mean the maximum of degrees of generators that appear in the relation.
For example, the degree of every relation of type $\text{R2}_{d,e}$ is $\max(d,e,d+e)$ 
 while the degree of a relation of type $\text{R3}^\bot_{d,e}$ or $\text{R3}^\angle_{d,e}$ is $\max(d,e)$.

@@

We denote by $Z_\alpha(A, M)$ the subgroup of $\overline{\St}(\Phi, A, M)$ generated by elements
 $x_{\beta}(m),\ \beta \in Z_+(\alpha)$ and $z_{\gamma}(m, \zeta),\ \gamma \in Z_0(\alpha),$ where $m \in M$, $\zeta \in A$.
It is not hard to see that \[Z_\alpha(A, M) = \Img\left(\overline{\St}(Z_0(\alpha), A, M) \to \overline{\St}(\Phi, A, M)\right) \rtimes \hat{U}(Z_+(\alpha), M). \]
It is clear, that $Z_\alpha(A, M)$ centralizes the root subgroup $X_\alpha(A)$ (cf. \cite[984]{St71}).

For the rest of this subsection $\Phi$ is a simply-laced root system of rank $\geq 3$ and $\alpha$ is a root of $\Phi$. Notice that in the simply-laced case the assumption $\alpha+\beta\not\in \Phi$ in the definition of $Z_0(\alpha)$ is superfluous, i.\,e. $Z_0(\alpha) = \{ \alpha\in\Phi \mid \alpha \perp \beta \}$
(cf.~\cite[Proposition~5.7]{St71}).

>>>>

\newcommand{\XX}{\mathcal{X}}           % (subset of) Steinberg generators
\newcommand{\RR}[1]{\mathcal{R}_{#1}}   % Steinberg relations
\newcommand{\catname}[1]{{\normalfont\textbf{#1}}} %Category name
\newcommand{\myol}[2][3]{{}\mkern#1mu\overline{\mkern-#1mu#2}}

@@

For an additive subgroup $A\subseteq R$ and $\alpha \in \Phi$ we denote by $X_\alpha(A)$ the corresponding {\it root subgroup} of $\St(\Phi, R)$, i.\,e. the subgroup generated by all $x_\alpha(a)$, $a \in A$.

@@

\subsection{Generators of relative Steinberg groups}
Denote by $\St(\Phi, I)$ the subgroup of $\St(\Phi, R)$ generated as a group by root unipotents of level $I$.
It is clear that $\overline{\St}(\Phi, R, I)$ contains $\St(\Phi, I)$ and, in fact, is its normal closure.
We also denote by $\myol{H}(\Phi, R, I)$ the subgroup of $\overline{\St}(\Phi, R, I)$ generated by the semisimple root elements $h_\alpha(u)$ and symbols $\{u, v\}$, $u \in (1+I)^\times$, $v \in R^\times$, $\alpha\in \Phi$.

@@

Recall that a closed root subset $S \subseteq \Phi$ is called {\it parabolic} (resp. {\it reductive}, resp. {\it special}) if $S \cup -S = \Phi$ (resp. $S = -S$, resp. $S \cap (-S) = \varnothing$).
The {\it special part} $\Sigma_S$ of a parabolic subset $S$, by definition, consists of all $\alpha \in S$ such that $-\alpha \not\in S$.
\begin{externaltheorem}[Stepanov] \label{thm:Stepanov} 
Let $S \subseteq \Phi$ be a parabolic subset of $\Phi$. Then the group $\overline{\St}(\Phi, R, I)$ is generated by the set $\mathcal{Z}(\Sigma_S, R, I)$.
 \end{externaltheorem} \begin{proof} See~\cite[Lemma~4]{S15}. \end{proof}

@@

\begin{lemma} \label{Crels}
The elements $c_\alpha(s, t)$ satisfy the following relations for all $s\in I,\ t\in J,\ \xi\in R$.
 \begin{enumerate}
 \item \label{C1} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = x_{\alpha}(- st\xi) \cdot x_{\alpha+\beta}(N_{\alpha,\beta}\cdot s^2t\xi)$ if $\alpha+\beta \in \Phi$;
 \item \label{C2} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = x_{\alpha}(st\xi + s^2t^2\xi) \cdot x_{\alpha-\beta}(N_{-\alpha, \beta}\cdot st^2\xi)$ if $\alpha-\beta \in \Phi$;  
 \item \label{C3} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = 1$ if $\alpha \perp \beta$;  
 \item \label{C4} If $\alpha+\beta\in\Phi$ then holds:
  \[c_{\alpha+\beta}(s, t\xi) = [x_{\beta}(st),\ x_{-\beta}(\xi)] ^ {x_{\alpha+\beta}(-s) x_{-\alpha}(\epsilon t)} \cdot c_{\alpha}(\epsilon s\xi, -\epsilon t)^{-1} \cdot x_{-\beta}(-st\xi^2),\]
  where $\epsilon = N_{\alpha,\beta}$.
 \end{enumerate}
\end{lemma}
\begin{proof}
The first assertion follows from the following computation:
\begin{multline} \nonumber [c_{\beta}(s, t),\ x_{\alpha}(\xi)]  = [x_{\beta}(s),\ x_{-\beta}(t)] \cdot [x_{-\beta}(t),\ x_{\beta}(s)\cdot x_{\alpha + \beta}(N_{\alpha, \beta} \cdot s\xi)] = \\
 = {}^{x_\beta(s)}\![x_{-\beta}(t),\ x_{\alpha+\beta}(N_{\alpha, \beta} \cdot s\xi)] = x_{\alpha}(- st\xi) \cdot x_{\alpha+\beta}(N_{\alpha,\beta} \cdot s^2t\xi). \end{multline}

For the proof of the other assertions we will need the following commutator identities:
\begin{align}
 [x, yz]^y =& [y^{-1}, x] \cdot [x, z], \label{rel43} \\
  [[x, y], z] =& \left([x^{-1}, [ y^{-1}, z]] ^ {y^{-1}} \cdot [y, [ z^{-1}, x^{-1}]] ^ {z^{-1}} \right)^{x^{-1}}. \label{HW-variant}
\end{align}

@@

 Now each factor $z_{\alpha_i}(f_i(0), \xi_i(X))$ can be written as follows:
 \[z_{\alpha_i}(f_i(0), \xi_i(0))^{x_{-\alpha_i}(X\xi'_i(X))} = [x_{-\alpha_i}(-X\xi'_i(X)),\ z_{\alpha_i}(f_i(0), \xi_i(0))] \cdot z_{\alpha_i}(f_i(0), \xi_i(0)).\]
 It follows from the formula $[g,\ h]^{h_1} = [h_1^{-1},\ g][g,\ h_1^{-1}h]$ that the subgroup \[C_0 := \left[\St(\Phi, XA[X]),\ \overline{\St}(\Phi, A, M)\right]\] is normalized by $\overline{\St}(\Phi, A, M)$. Thus, we conclude that $g_1(X)$ is congruent to $\prod_i z_{\alpha_i}(f_i(0), \xi_i(0)) = g(0) = 1$ modulo $C_0$,
 which implies the assertion. \qedhere \end{proof}

For a closed root subset $S \subseteq \Phi$ we denote by $U(S, M)$ the subgroup of $\St(\Phi, A)$ generated by root subgroups $X_\alpha(M)$ corresponding to all $\alpha \in S$. We denote by $\Phi^+$ (resp. $\Phi^-$) the subsets of positive (resp. negative) roots of $\Phi$ with respect to some chosen order on $\Phi$.

\begin{externaltheorem}[Stein] \label{thm:Stein} One has $\overline{\St}(\Phi, A, M) = U(\Phi^+, M) \cdot \myol{H}(\Phi, A, M) \cdot U(\Phi^-, M).$ \end{externaltheorem} \begin{proof} See~\cite[Theorem~2.4]{Ste73}. \end{proof}

\begin{prop} \label{Kgen} The subgroup $K(A[X], M[X])$ is generated as an abstract group by the subgroup $\overline{\St}(\Phi, A[X], XM[X])$ and
 the elements $[x_\alpha(m), x_{-\alpha}(X\xi)]$, $m \in M$, $\xi \in A[X]$, $\alpha \in \Phi$. \end{prop}
\begin{proof} From~\eqref{eq:conj-h-x} we obtain that $\myol{H}(\Phi, A, M)$ normalizes both $\St(\Phi, XA[X])$ and $\St(\Phi, M)$ and, moreover, that $[\myol{H}(\Phi, A, M),\ \St(\Phi, XA[X])] \subseteq \overline{\St}(\Phi, A[X], XM[X])$.

Denote by $C_1$ the commutator subgroup $[\St(\Phi, XA[X]),\ \St(\Phi, M)]$.
It is clear that for $g \in \St(\Phi, XA[X])$, $h \in \myol{H}(\Phi, A, M)$, $u^+ \in U(\Phi^+, M)$, $u^- \in U(\Phi^-, M)$ holds:
\[ [g,\ h u^+ u^-] = [g,\ h] \cdot [{}^{h}\!g,\ {}^{h}\!(u^+u^-)] \in \St(\Phi, A[X], XM[X]) \cdot C_1.\]
Since $C_0$ is generated by the above commutators and $\overline{\St}(\Phi, A[X], XM[X])$ is a normal subgroup of $\St(\Phi, A[X])$
we obtain that $C_0 \subseteq \overline{\St}(\Phi, A[X], XM[X]) \cdot C_1$ and consequently that
$K(A[X], M[X]) = \overline{\St}(\Phi, A[X], XM[X]) \cdot C_1.$

@@

\section{Proof of the main result}
The main result of this section is~\cref{thm:P1glueing}, which is a direct generalization of~\cite[Proposition~4.3]{Tu83}. 
The object playing a key role in its proof is a certain action of the group $G = \St(\Phi, A[X\inv] + M[X])$ on a certain set $\overline{V}$, which is defined in~\cref{sec:V-construction}. Later, we will see that $\overline{V}$ is, in fact, a set-theoretic $G$-torsor.
To be able to write an explicit formula for this action we need two major ingredients. The first one is ~\cref{lemma33}, which gives a presentation of $G$ with much fewer generations and relations than in the original presentation~\eqref{rel:add}-\eqref{rel:CCF}. The other ingredients are certain subgroups $P_\alpha(0)$, $P_\alpha(*)$ of $\St(\Phi, A[X, X\inv])$ modeled after the nameless groups from~\cite[Lemma~3.4]{Tu83}. %The definition and properties of these groups are given in Sections~\ref{sec:Pa0-basic}--\ref{sec:Pa0-Steinberg}.

@@

Whenever the coefficient $\xi$ of a Steinberg generator $g = x_\alpha(\xi)$ of $\St(\Phi, B)$ is a homogeneous element of $B$, i.\,e. $\xi \in B_d$ for some $d \in \mathbb{Z}$,
 we call the corresponding generator $g$ {\it homogeneous of degree $d$}.
It is not hard to show that $\St(\Phi, B)$ can be presented by the set of all homogeneous Steinberg generators modulo the following set of Steinberg relations
(below $a, a' \in B_d$, $b\in R_e$ and $d,e \in \mathbb{Z}$): 
\begin{align}
&\,\,x_{\alpha}(a)\cdot x_{\alpha}(a') =  x_{\alpha}(a+a'),                        & \tag{R$1_d$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= x_{\alpha+\beta}(N_{\alpha, \beta} \cdot ab),   & \alpha + \beta \in \Phi,\ \tag{R$2_{d,e}$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= 1,                                              & \alpha - \beta \in \Phi,\ \tag{R$3^\angle_{d,e}$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= 1,                                              & \alpha \perp \beta.\ \tag{R$3^\bot_{d,e}$}
\end{align}
By the degree of a Steinberg relation we mean the maximum of degrees of generators that appear in the relation.
For example, the degree of every relation of type $\text{R2}_{d,e}$ is $\max(d,e,d+e)$ 
 while the degree of a relation of type $\text{R3}^\bot_{d,e}$ or $\text{R3}^\angle_{d,e}$ is $\max(d,e)$.

@@

We denote by $Z_\alpha(A, M)$ the subgroup of $\overline{\St}(\Phi, A, M)$ generated by elements
 $x_{\beta}(m),\ \beta \in Z_+(\alpha)$ and $z_{\gamma}(m, \zeta),\ \gamma \in Z_0(\alpha),$ where $m \in M$, $\zeta \in A$.
It is not hard to see that \[Z_\alpha(A, M) = \Img\left(\overline{\St}(Z_0(\alpha), A, M) \to \overline{\St}(\Phi, A, M)\right) \rtimes U(Z_+(\alpha), M). \]
It is clear, that $Z_\alpha(A, M)$ centralizes the root subgroup $X_\alpha(A)$ (cf. \cite[984]{St71}).

For the rest of this subsection $\Phi$ is a simply-laced root system of rank $\geq 3$ and $\alpha$ is a fixed root of $\Phi$. Notice that in the simply-laced case the assumption $\alpha+\beta\not\in \Phi$ in the definition of $Z_0(\alpha)$ is superfluous, i.\,e. $Z_0(\alpha) = \{ \alpha\in\Phi \mid \alpha \perp \beta \}$
(cf.~\cite[Proposition~5.7]{St71}).
